
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  EGSnrc manual: associated documents
%  Copyright (C) 2015 National Research Council Canada
%
%  This file is part of EGSnrc.
%
%  EGSnrc is free software: you can redistribute it and/or modify it under
%  the terms of the GNU Affero General Public License as published by the
%  Free Software Foundation, either version 3 of the License, or (at your
%  option) any later version.
%
%  EGSnrc is distributed in the hope that it will be useful, but WITHOUT ANY
%  WARRANTY; without even the implied warranty of MERCHANTABILITY or FITNESS
%  FOR A PARTICULAR PURPOSE.  See the GNU Affero General Public License for
%  more details.
%
%  You should have received a copy of the GNU Affero General Public License
%  along with EGSnrc. If not, see <http://www.gnu.org/licenses/>.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%  Author:          Iwan Kawrakow, 2003
%
%  Contributors:    Blake Walters
%                   Frederic Tessier
%                   Ernesto Mainegra-Hing
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Replace commented line for the one with fixed date when commiting
% Beware: Using the macro below conflicts between CVS and latex!!!
% \lfoot[{\sffamily {\leftmark}}]{{\small Last edited $Date: 2011/05/02 18:36:27 $
\lfoot[{\sffamily {\leftmark}}]{{\small Last edited 2011/05/02 18:30:41
}}

\index{associated documents}
There are a variety of papers which have been written about EGSnrc and
several NRC reports which are part of the distribution.  These are
listed below.  There are also a large number of papers which have been
written about EGS4.

\subsubsection{Refereed Papers}
\begin{itemize}
\item {\bfseries Accurate condensed history Monte Carlo simulation of
     electron transport. I. EGSnrc, the new EGS4 version:} \\
 I. Kawrakow, Medical Physics, {\bf 27} (2000) 485 -- 498.\\
Describes the overall implementation of the new electron transport
physics in EGSnrc.

\item {\bfseries Accurate condensed history Monte Carlo simulation of
     electron transport. II. Application to ion chamber response
     simulations:}\\
 I. Kawrakow, Medical Physics, {\bf 27} (2000) 499 -- 513.\\
Quantifies which problems in EGS4 led to the difficulties simulating ion
chamber response accurately and demonstrates that EGSnrc does not have
these problems.

\item {\bfseries Monte Carlo study of Spencer-Attix cavity theory at low
     photon energies:}\\
J. Borg, I. Kawrakow, D.W.O. Rogers and J.P.  Seuntjens, Med. Phys. {\bf
 27} (2000) 1804 -- 1813.
\\Uses code to explore the accuracy of Spencer-Attix cavity theory.
Paper demonstrates the accuracy of the EGSnrc code system for calculations
related to real ion chambers.


\item {\bfseries On the representation of electron multiple
elastic-scattering distributions for Monte Carlo calculations:}\\
I. Kawrakow and A. F.  Bielajew, Nucl. Inst. Meth. B 134, (1998) 325 --
336\\ Describes the multiple scattering theory used in EGSnrc.

\index{Bielajew, Alex}
\item {\bfseries On the condensed history technique for electron
transport:}\\
I. Kawrakow and A. F.  Bielajew, Nuclear Instruments and
Methods B 142 (1998) 253 -- 280.\\ Describes the electron transport
algorithm used in EGSnrc and demonstrates its improved accuracy compared
to all other published algorithms.


\end{itemize}
\subsubsection{Internal Reports}
\begin{itemize} 
\item {\bfseries Monte Carlo calculated wall and axial non-uniformity
     corrections for primary standards of air kerma},\\
D. W. O.  Rogers and J. Treurniet, NRC Report PIRS--663, May 1999.\\Describes extensive EGSnrc
calculations of ion chamber response and comparison to experimental data
from standards labs of response vs wall thickness. Also notes that results
for these calculations, which are or correction factors, are the same as
for EGS4/PRESTA calculations.
\end{itemize} 

\subsubsection{Manuals etc}

\begin{itemize}
\item {\bfseries NRC User Codes for EGSnrc}\\ D.W.O. Rogers, I. Kawrakow, J.P.
Seuntjens and B.R.B. Walters, NRC Report PIRS--702, May 2009.\\
Describes the EGSnrc user codes DOSRZnrc, CAVRZnrc, FLURZnrc and
SPRRZnrc as well as the generalized input routine developed for use with 
EGSnrc user codes.

\item {\bfseries EGS\_Windows4.0 User's Manual},\\ J. A.  Treurniet and D.
W. O.  Rogers, NRC Report PIRS--669, Oct 1999.\\
Describes the latest version of  EGS\_Windows which works on any X-windows
based system and can be used to display EGSnrc histories in 3-D.\\


\item {\bfseries  QA tests and comparisons of the EGSnrc system
with EGS4},\\ B. R. B. Walters, J. Treurniet, D. W. O. Rogers
               and I. Kawrakow, NRC Report PIRS-703, March 2000.\\
Summarizes a large number of comparisons between EGS4 and EGSnrc to
highlight some differences and similarities.\\


\item {\bfseries  EGSnrcMP: the multi-platform environment for EGSnrc}, \\
I. Kawrakow, E. Mainegra-Hing and D. W. O. Rogers
           NRC Report PIRS-877, Sept 2006.\\
Describes the changes in the system of scripts used to run the EGSnrc
system. These major changes mean that EGSnrc now works under the Windows OS
and there are GUI's for installing and running the system.\\


\item {\bfseries egs\_inprz, a GUI for the NRC RZ user-codes}, \\
E. Mainegra-Hing, NRC Report PIRS-801, May 2005.\\
Manual for GUI for RZ user codes.\\


\item {\bfseries EGSnrc C++ class library},\\
I. Kawrakow, NRC Report PIRS-898, Apr 2006.\\
Manual describing the geometry and source modules available with the
EGSnrc C++ class library.  Also describes how to build a user code using
this library.  Includes descriptions of the C++ user codes, {\tt cavity},
{\tt egs\_chamber}, {\tt egs\_cbct}, {\tt egs\_fac} and {\tt egs\_fac}.\\

\end{itemize}


